---
title: "Answers to exercises Session 5"
author: "Marjolein Fokkema"
output: pdf_document
---

# Exercise 1

Read in data:

```{r}
load("MASQ.Rda")
set.seed(1)
train <- sample(1:nrow(MASQ), size = nrow(MASQ)*.8)
summary(MASQ)
```

Fit a smoothing spline of the `AD` variable to predict `D_DEPDYS`:

```{r, warning=FALSE, message=FALSE}
library("mgcv")
GAM <- gam(D_DEPDYS ~ s(AD, bs = "cr"), 
           data = MASQ[train, ], method = "REML", family = "binomial")
summary(GAM)
plot(GAM, residuals = TRUE)
``` 

Inspect the basis functions that were created for `AD`: 

```{r}
mod_mat <- model.matrix(GAM)
matplot(sort(MASQ$AD[train]), mod_mat[order(MASQ$AD[train]), ], type = "l",
        xlab = "AD", ylab = "Basis function")
```




# Exercise 2: Multiple predictors

```{r, fig.width=7, fig.height=4, warning=FALSE, message=FALSE}
library("mgcv")
GAM <- gam(D_DEPDYS ~ s(AD) + s(AA) + s(GDD) + s(GDA) + s(GDM) + s(leeftijd) + geslacht, 
           data = MASQ[train, ], method = "REML", family = "binomial")
summary(GAM)
par(mfrow = c(2, 3))
plot(GAM)
```

We compute the mean squared error and misclassification rate using predicted probabilities, for both training and test observations:

```{r}
y_train <- as.numeric(MASQ[train, "D_DEPDYS"]) - 1
y_test <- as.numeric(MASQ[-train, "D_DEPDYS"]) - 1

## Training data
GAM_preds_train <- predict(GAM, newdata = MASQ[train, ], type = "response")
mean((y_train - GAM_preds_train)^2) ## Brier score
tab_train <- prop.table(table(MASQ[train, "D_DEPDYS"], GAM_preds_train > .5)) ## confusion matrix
1 - sum(diag(tab_train)) ## MCR

## Test data
GAM_preds_test <- predict(GAM, newdata = MASQ[-train, ], type = "response")
mean((y_test - GAM_preds_test)^2) ## Brier score
tab_test <- prop.table(table(MASQ[-train, "D_DEPDYS"], GAM_preds_test > .5)) ## confusion matrix
1 - sum(diag(tab_test)) ## MCR
```

The Brier score and confusion matrices are quite similar between training and test data, indicating little overfitting. 

```{r, warning=FALSE, message=FALSE}
## Or, compute ROC curve
library("pROC")
auc(resp = y_test, pred = GAM_preds_test)
```